How to build an homogeneous language for heterogeneous platforms

### 2013-05-31 papers/gpce2013/gpce2013.tex
abstract + intro

\begin{abstract}
Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
and abstracting over the clients and servers differences without trading this engineering comfort for performance?
This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
code leveraging the target platform characteristics, (2) be shared between client and server code. Our DSL gives a
performance / expressiveness ratio about two times higher than other approaches.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

@@

The more interactive the application is, the more logic needs to be duplicated between the server-side
and the client-side (explain why?), and the higher is the amount of client-side code. Developers use libraries and
frameworks to get high-level abstractions on client-side, making their code easier to reason about and to
maintain, but also making their code run less efficiently (abstraction penalty).

@@

\texttt{js-scala} is such a compiled embedded DSL defined in Scala that generates JavaScript code, making it possible
to write the client-side code of Web applications using JavaScript~\cite{Kossakowski12_JsDESL}. This paper
enriches js-scala in order to solve the problems described above. We use staging (explain?) to generate efficient
code for typical abstractions used for Web programming and to generate efficient specialized code for both client and
server sides for shared abstractions. More precisely, we demonstrate the following features:

\begin{itemize}
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs;
 \item Usage of monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side;
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic.
\end{itemize}

Compared to libraries offering the same level of abstraction, we observe that js-scala produces significantly smaller
executables with better runtime performances.

The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
cross-compiling languages. Section \ref{contribution} presents our contribution. Section \ref{validation} evaluates
our contribution. Section \ref{discussion} concludes.

\section{Related Work (?)}

@@

Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
language (this is also referred to as \emph{embedded} domain specific languages). Figure \ref{langs} (c) depicts this
approach. The general purpose language is used as a host language and does not need to be modified if a new concept
is introduced, because concepts are defined as pure libraries. However, this approach gives no opportunity to
translate a concept efficiently according to the host platform characteristics. Examples of languages following this
approach are Java/GWT, Kotlin, HaXe and SharpKit.

@@

\section{High-Level Abstractions For Client-Side Code}
\label{contribution}

@@

\section{High-Level Abstractions Shared on Clients and Servers}

>>>>

\begin{abstract}
Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
and abstracting over the clients and servers differences without trading this engineering comfort for performance?
This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
code leveraging the target platform characteristics, (2) be shared between client and server code. Our DSLs have a
performance / expressiveness ratio about two times higher than other approaches.
\end{abstract}

\category{D.3.3}{Programming Languages}{Language Constructs and Features}

\terms Languages, Software Engineering

\keywords Heterogeneous Code Generation, Domain-specific languages

@@

The more interactive the application is, the more logic needs to be duplicated between the server-side
and the client-side, and the higher is the amount of client-side code. Developers use libraries and frameworks to get
high-level abstractions on client-side, making their code easier to reason about and to maintain, but also making
their code run less efficiently (abstraction penalty).

@@

\texttt{js-scala} is such a compiled embedded DSL defined in Scala that generates JavaScript code, making it possible
to write the client-side code of Web applications using JavaScript~\cite{Kossakowski12_JsDESL}. This paper
enriches \texttt{js-scala} in order to solve the problem described above. We use staging~\cite{Jorring1986_Staging}
to:

\begin{itemize}
 \item generate efficient code for typical abstractions used in Web programming;
 \item generate specialized code for both client and server sides for shared abstractions.
\end{itemize}

Though the code written in \texttt{js-scala} is high-level and can be shared between clients and
servers, it has the same runtime performances as hand-tune low-level JavaScript code. We observed a performance /
expressiveness ratio two times higher than using other approaches.

The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
cross-compiling languages. Sections \ref{contribution} and \ref{contribution-shared} present our contribution.
Section \ref{implementation} gives implementation details. Section \ref{validation} evaluates the contribution.
Section \ref{discussion} concludes.

\section{Background}

@@

Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
language (this is also referred to as \emph{embedded} domain specific languages). Figure \ref{langs} (c) depicts this
approach. The general purpose language is used as a host language and does not need to be modified if a new concept
is introduced, because concepts are defined as pure libraries. However, this approach gives no opportunity to
translate a concept efficiently according to the target platform characteristics. Examples of languages following
this approach are Java/GWT, Kotlin, HaXe and SharpKit.

@@

\section{High-Level Abstractions for Client-Side Code}
\label{contribution}

@@

\section{High-Level Abstractions Shared on Clients and Servers}
\label{contribution-shared}
